---
title: "Monday Analysis"
authors: "Adeyemi Adepetun/Jiashu Zhao"
date: "7/4/2021"
output: rmarkdown::github_document
params:
  day: "Mon"
---

# Packages required to load and analyse the data
The following packages are needed for reading and analyzing the data:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
packages <- c("readr","ggplot2","dplyr","knitr", "caret", "tidyr","lubridate")
lapply(packages, library, character.only = TRUE)
```

# `r toupper(params$day)` Report

## Introduction







Reading in the data using a relative path
```{r}
day_data  <-  read_csv("day.csv")
```
Converting the weekday variable to a factor containing the 7 days of the week:

```{r}
day_as_fac_data <- day_data %>%
    mutate(weekday_alt = wday(weekday +1,label=TRUE)) 
day_as_char_data <- day_as_fac_data %>% mutate_if(is.factor, as.character)
```

Setting up the dataset for later automation. The code will be automated to analysis data for the day from the day: "weekday" seen in the YAML header:
```{r}
data_weekday <- day_as_char_data %>% 
  filter(day_as_char_data$weekday_alt == params$day)
```



## Train/Test split
Doing a 70:30 split for training/testing:
```{r}
set.seed(365)
train  <-  sample(1:nrow(data_weekday), size = 0.7*nrow(data_weekday))
test  <- setdiff(1:nrow(data_weekday), train)
day_data_Train <- data_weekday[train, ]
day_data_Test <- data_weekday[test, ]
```

