---
title: "Monday Analysis"
authors: "Adeyemi Adepetun/Jiashu Zhao"
date: "7/4/2021"
output: 
  rmarkdown::github_document:
    toc: yes
params:
  day: "Mon"
---

```{r global_options}
knitr::opts_chunk$set(fig.path='Figs/')
```


# Packages required to load and analyse the data
The following packages are needed for reading and analyzing the data:
```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
packages <- c("readr","ggplot2","dplyr","knitr", "caret", "tidyr","lubridate", "forcats")
lapply(packages, library, character.only = TRUE)
```

# `r toupper(params$day)` Report

## Introduction

In this project we decided to present an analysis report about bike sharing data. The data comes from the UCI Machine Learning Repository, contains the hourly and daily count of rental bikes between years 2011 and 2012 in Capital bikeshare system with the corresponding weather and seasonal information. 731 days of data were collected.

We would like to explore if the weather and season factors may affect the number of users. We aim to develop accurate models by both linear regression and ensemble tree.

The following 13 characteristics were collected for each days along with the count of casual/registered users.

`Date`

`Season`: Winter, spring, summer and fall.

`Year`: 2011 and 2012.

`Month`

`Hour`

`Holiday`: Weather day is holiday or not.

`Weekday`

`workingday`: if day is neither weekend nor holiday is 1, otherwise is 0.

`Weathersit`: 4 levels from smooth to terrible.

`Temperature`: Normalized.

`Feeling temperature`: Normalized.

`Humidity`: Normalized.

`Windspeed`: Normalized.


## Data preparation

Reading in the data using a relative path

```{r}
day_data  <-  read_csv("day.csv")
```

Converting the year, season, and weather situation variables:

```{r}
#categorizing the season variable as factor
day_data <- mutate(day_data, Season = 
                  ifelse(season == 1, "Winter",
                  ifelse(season == 2, "Spring",
                  ifelse(season == 3, "Summer",
                  ifelse(season == 4, "Fall", "")))))
day_data$Season <- factor(day_data$Season, levels=c("Winter","Spring", "Summer","Fall"))

#categorizing the year variable as factor
day_data <- mutate(day_data, Year = 
                  ifelse(yr == 0, "2011",
                  ifelse(yr == 1, "2012", "")))
day_data$Year <- factor(day_data$Year, levels=c("2011","2012"))

#categorizing the weathersit variable as factor
day_data <- mutate(day_data, weather_situation = 
                  ifelse(weathersit == 1, "Clear Weather",
                  ifelse(weathersit == 2, "Misty/Cloudy Weather",
                  ifelse(weathersit == 3, "Light Snow/Rain/Thunderstorm Weather",
                  ifelse(weathersit == 4, "Heavy Rain/Snow/Fog/Icy Weather", "")))))
day_data$weather_situation <- factor(day_data$weather_situation, levels=c("Clear Weather","Misty/Cloudy Weather", "Light Snow/Rain/Thunderstorm Weather","Heavy Rain/Snow/Fog/Icy Weather"))

#reordering the day_data and renaming mth, cnt and hum
day_data <- day_data %>% select(-c("instant","season","yr", "weathersit")) %>% rename(month = mnth,count = cnt,humidity = hum)
```

Converting the weekday variable to a factor containing the 7 days of the week:

```{r}
day_as_fac_data <- day_data %>%
    mutate(week_day = wday(weekday +1,label=TRUE,locale="English_United States")) 
day_as_char_data <- day_as_fac_data %>% mutate_if(is.factor, as.character)
```


Setting up the dataset for later automation. The code will be automated to analysis data for the day from the day: "weekday" seen in the YAML header:
```{r}
data_weekday <- day_as_char_data %>% 
  filter(day_as_char_data$week_day == params$day)
```

## Train/Test split
Doing a 70:30 split for training/testing:
```{r}
set.seed(365)
train  <-  sample(1:nrow(data_weekday), size = 0.7*nrow(data_weekday))
test  <- setdiff(1:nrow(data_weekday), train)
day_data_Train <- data_weekday[train, ]
day_data_Test <- data_weekday[test, ]
```

## Data Summarizations and discussions

### Overall Count

The averages for selected variables for each year is shown below

```{r}
grouped_mean  <-  day_data_Train %>%
  group_by(Year) %>%
  summarise(
    n=n(),
    avg_registered_user = round(mean(registered), 0),
    avg_casual_user = round(mean(casual), 0),
    avg_windspeed = round(mean(windspeed*67), 2),
    avg_humidity = round(mean(humidity*100), 2),
  )

kable(grouped_mean, caption="Averages of selected variables for each year")
```

The standard deviations by year for each of the variables above are summarized below.
```{r}
grouped_sd  <-  day_data_Train %>%
  group_by(Year) %>%
  summarise(
    n=n(),
    avg_registered_user = round(sd(registered), 0),
    avg_casual_user = round(sd(casual), 0),
    avg_windspeed = round(sd(windspeed*67), 3),
    avg_humidity = round(sd(humidity*100), 3),
  )
  
kable(grouped_sd, caption="Standard deviation of selected variables for each year")
```

### Weather Condition

The table for weather condition vs. year for this particular day is shown below:
```{r}
year_season_table <- table(day_data_Train$Year, day_data_Train$weather_situation )
kable(year_season_table, caption="Table of Year vs. Weather situation")
```

This is visualized below:


```{r}
ggplot(data=day_data_Train, aes(x=day_data_Train$Year))+
  geom_bar(aes(fill=as.factor(day_data_Train$weather_situation)))+ 
  labs(x="Year", title="Weather situation by year for this week day")+ 
  scale_fill_discrete(name="Weather situation", labels=c("Clear Weather","Misty/Cloudy Weather", "Light Snow/Rain/Thunderstorm Weather","Heavy Rain/Snow/Fog/Icy Weather")) 
```

This bar plot shows the weather pattern for this particular day in 2011 and 2012.

We can also see if for a given year, weather condition has an effect on the number of rentals for this week day

```{r}
ggplot(data=day_data_Train, aes(x=weather_situation, y= count))+
  geom_boxplot(aes(fill=weather_situation))+
  labs(x="Weather Situation", y="Total number of rentals", fill="Weather Situation", title = "Boxplot of Total number of rentals by Weather condition for this week day")+  theme(axis.text.x = element_text(size  = 10,angle = 45,hjust = 1,vjust = 1)) 
```

### Seasons

The number of casual and registered users on this particular day from 2011-2012 are categorized below:
```{r}
#categorizing the number casual users as factor
day_data_Train <- mutate(day_data_Train, casual_format = 
                  ifelse(casual %in% 0:1000, "Low number of casual users",
                  ifelse(casual %in% 1000:2000, "Medium number of casual users",
                  ifelse(casual > 2000, "High number of casual users", ""))))
day_data_Train$casual_format <- factor(day_data_Train$casual_format, levels=c("Low number of casual users","Medium number of casual users", "High number of casual users"))

#categorizing the number registered users as factor
day_data_Train <- mutate(day_data_Train, registered_format = 
                  ifelse(registered %in% 0:2000, "Low number of registered users",
                  ifelse(registered %in% 2000:4000, "Medium number of registered users",
                  ifelse(registered > 4000, "High number of registered users", ""))))
day_data_Train$registered_format <- factor(day_data_Train$registered_format, levels=c("Low number of registered users","Medium number of registered users", "High number of registered users"))
```

The number of casual users vs season for 2011-2012 are shown in the table and bar chart below:

```{r}
casual_users_table <- table(day_data_Train$casual_format, day_data_Train$Season )
kable(casual_users_table, caption="Casual users vs Season for 2011-2012")

ggplot(data=day_data_Train, aes(x=day_data_Train$Season))+
  geom_bar(aes(fill=as.factor(day_data_Train$casual_format)))+ 
  labs(x="Season", title="Casual users vs Season for this week day from 2011-2012")+ 
  scale_fill_discrete(name="Casual Users", labels=c("Low number of casual users","Medium number of casual users", "High number of casual users")) + 
  coord_flip()
```

The same information for  number of registered users is shown in the table and bar chart below:

```{r}
registered_users_table <- table(day_data_Train$registered_format, day_data_Train$Season )
kable(registered_users_table, caption="Registered users vs Season for 2011-2012")

ggplot(data=day_data_Train, aes(x=day_data_Train$Season))+
  geom_bar(aes(fill=as.factor(day_data_Train$registered_format)))+ 
  labs(x="Season", title="Registered users vs Season for this week day from 2011-2012")+ 
  scale_fill_discrete(name="Registered Users", labels=c("Low number of registered users","Medium number of registered users", "High number of registered users")) + 
  coord_flip()
```

We can inspect the trend of casual and registered users across seasons to see if there is a seasonal effect present.

Next, it would be nice to see how total bike rentals vary by season and weather situation for this particular day.
```{r}
ggplot(data=day_data_Train, aes(x=Season, y= count))+
  geom_boxplot(aes(fill=Season))+
  labs(x="Season", y="Total number of rentals", fill="Season", title = "Boxplot of Total number of rentals by Season for this week day")
```

This plot will let us know if there is a seasonal effect on number of rentals for this particular week day.

### Temperature and Feeling Temperature

We are curious about how difference is between temperature and feeling temperature, so we decided to make a comparison line plot.

```{r}
tem_plot <- day_data_Train %>%
  select(dteday, temp, atemp, Year, count) %>%
  gather('temp','value',2:3)

ggplot(tem_plot) + geom_line(aes(x=dteday,y=value,color=temp)) + facet_wrap(~Year, scales = 'free') + theme_bw() + scale_colour_discrete(name = '', labels=c('temp','feeling temp')) + labs(x='Date', y='Temp', title = 'Temperature vs. feeling temperature')
```

Basically, When the temperature is high, the body temperature will be lower; when the actual temperature is very low, the opposite is true.

There's no difference between these 2 variables, so we decided to drop `feeling temperature` due to the consideration of collinearity. Besides, the variance of `temperature` is a little bit larger than `feeling temperature`, so we expect `temperature` to be more sensitive when fitting the model.

### Humidity

We would like to explore the general relationship between humidity and users count.

```{r}
hum_plot <- day_data_Train %>%
  select(humidity, casual, registered, Year) %>%
  gather('regist', 'count', 2:3)

ggplot(data = hum_plot, aes(x=humidity, y=count, group=regist)) + geom_point(aes(color=regist)) + geom_smooth(aes(group=regist), color='black') + facet_wrap(~Year, scales = 'free') + theme_bw() + scale_colour_discrete(name = '') + labs(title = 'Relationship between Humidity and User Count')
```

We can inspect the relationship of users with humidity using this plot roughly. The points are nearly evenly distributed, which means humidity can seldom affect the users if we control all other variables.

### Wind Speed

```{r}
ggplot(data = day_data_Train, aes(x=windspeed, y=..density..)) + geom_histogram(bins = 20) + geom_density(size=2.5,color='red',adjust=1/3) + facet_wrap(~Year, scales = 'free') + theme_bw()
```

This plot shows the distribution of windspeed. Although the number of samples is small, it still shows a certain degree of normality.

Same as humidity, we are curious about the relationship between windspeed and user count after controlled all other variables.

```{r}
wind_plot <- day_data_Train %>%
  select(windspeed, casual, registered, Year) %>%
  gather('regist', 'count', 2:3)

ggplot(data = wind_plot, aes(x=windspeed, y=count, group=regist)) + geom_point(aes(color=regist)) + geom_smooth(aes(group=regist), color='black') + facet_wrap(~Year, scales = 'free') + theme_bw() + scale_colour_discrete(name = '') + labs(title = 'Relationship between Windspeed and User Count')
```

We can inspect the relationship of users with windspeed using this plot roughly.


## Modeling

### Linear Regression Model
Linear regression modeling is one of the supervised learning methods where the output is known, and the goal is to establish a function that best approximates the relationship between desired outputs and the provided sample data. Specifically, linear regression accomplishes this by learning a model that best fits the linear relationship between the predictor and response variables.The model is fit by minimizing the sum of squared residuals (difference between the observed and predicted responses). Linear regression models fall into two categories: Simple Linear Regression and Multiple Linear Regression.

Simple Linear Regression Model is one where there is only one independent variable (or predictor) and the goal is to learn the linear relationship between it and the response variable. Multiple Linear Regression (MLR) has more than one predictor, and the although relationship between predictors and response remains linear in terms of the model parameters, the MLR model could contain interaction, quadratic and polynomial terms.

Underlying assumptions for the linear regression model are 
* Linearity: The model is linear in model parameters (betas) (Can be checked using histogram or Q-Q plots) 
* Normality: The predictor and response variables are multivariate normal  
* Multicollinearity: There is little to no multicollinearity among the predictor variables. (can be checked using Variance Inflation Factor)
* Homoscedasticity: Residuals are randomly distributed across the regression line (Can be checked using the Residual vs. Fitted value scatter plot. The plot must have to discernable pattern)  
* Autocorrelation: Residuals must be independent of each other (Can be checked using Durbin-Watson's test).
